### LiverMA.Rmd
### E Flynn
### Created: March 2018
### Last Updated: November 2018


```{r, message=FALSE, warning=FALSE}
require('MetaIntegrator')
require('miceadds')
require('reshape2')
require('biomaRt')
require('gridExtra')
require('gplots')

source("processing_utils.R")
```

Load the data
```{r}
resListMI <- lapply(sapply(c(1:16), function(x) paste(x, "g", sep="")), extractStudyMI)

names(resListMI) <- sapply(resListMI, function(x) x$formattedName)

```



Divide into discovery + validation
- need only 300 for discovery
- leave 9588
- try leaveOneOut
```{r}
discoveryStudies <- c(1:7, 9:11) # 434
validationStudies <- c(8, 12:16) # put 8 in validation because it is one of the largest!

metaObj <- list()
metaObj$originalData <- resListMI[names(resListMI)[discoveryStudies]]
checkDataObject(metaObj, "Meta", "Pre-Analysis")

metaObjv <- list()
metaObjv$originalData <- resListMI[names(resListMI)[validationStudies]]
checkDataObject(metaObjv, "Meta", "Pre-Analysis")


metaObj <- runMetaAnalysis(metaObj)
metaObjv <- runMetaAnalysis(metaObjv)
save(metaObj, metaObjv, file="data/results/metaResults_0925.RData")

```


Check Distribution of Effect Sizes
(Separated out + zoomed in for clarity!)
```{r}
ggplot(melt(metaObj$metaAnalysis$datasetEffectSizes[,1:3], 
        varnames = c("Gene", "Study")), aes_string(x = "value", 
        colour = "Study")) + geom_density(size = 1.1) + theme_bw() + 
        scale_color_discrete(name = "Dataset") + xlim(-3,3) # concerned about GSE38941... Wider than normal
ggplot(melt(metaObj$metaAnalysis$datasetEffectSizes[,4:6], 
        varnames = c("Gene", "Study")), aes_string(x = "value", 
        colour = "Study")) + geom_density(size = 1.1) + theme_bw() + 
        scale_color_discrete(name = "Dataset")+ xlim(-3,3)
ggplot(melt(metaObj$metaAnalysis$datasetEffectSizes[,7:10], 
        varnames = c("Gene", "Study")), aes_string(x = "value", 
        colour = "Study")) + geom_density(size = 1.1) + theme_bw() + 
        scale_color_discrete(name = "Dataset")+ xlim(-3,3)
```

```{r}
ggplot(melt(metaObjv$metaAnalysis$datasetEffectSizes[,1:3], 
        varnames = c("Gene", "Study")), aes_string(x = "value", 
        colour = "Study")) + geom_density(size = 1.1) + theme_bw() + 
        scale_color_discrete(name = "Dataset")+ xlim(-3,3) # GSE9588 looks narrow, other two look a little wide...
ggplot(melt(metaObjv$metaAnalysis$datasetEffectSizes[,4:6], 
        varnames = c("Gene", "Study")), aes_string(x = "value", 
        colour = "Study")) + geom_density(size = 1.1) + theme_bw() + 
        scale_color_discrete(name = "Dataset")+ xlim(-3,3) # E-MEXP-3291 looks a little wide
```
Most of the effect sizes look mostly normal and symmetric, but a few are a little wide, and one is a little narrow. 


Filter the resulting genes by FDR and and effect size. 
```{r}
metaObj <- filterGenes(metaObj, isLeaveOneOut = TRUE, effectSizeThresh=0.3, FDRThresh = 0.2, numberStudiesThresh = 2)
metaObj <- filterGenes(metaObj, isLeaveOneOut = FALSE, effectSizeThresh=0.4, FDRThresh = 0.05, numberStudiesThresh = 2)
metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0$posGeneNames # up in males
metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0$negGeneNames # up in females

metaRes <- summarizeFilterResults(metaObj, "FDR0.2_es0.3_nStudies2_looaTRUE_hetero0")
save(metaRes, file="data/results/metaSig_1107.RData")
```

Check the separation of classes in the validation - it looks good (this is to be expected).
```{r}
filtRes <-metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0
p1 <- violinPlot(filtRes, metaObjv$originalData$GSE9588, labelColumn = 'sexLabels')
p2 <- violinPlot(filtRes, metaObjv$originalData$GSE33814, labelColumn = 'sexLabels')
p3 <- violinPlot(filtRes, metaObjv$originalData$GSE48452, labelColumn = 'sexLabels')
p4 <- violinPlot(filtRes, metaObjv$originalData$E.MEXP.3291, labelColumn = 'sexLabels')
p5 <- violinPlot(filtRes, metaObjv$originalData$GSE25935, labelColumn = 'sexLabels')
p6 <- violinPlot(filtRes, metaObjv$originalData$GSE28893, labelColumn = 'sexLabels')

grid.arrange(p1, p2, p3, p4, p5, p6, nrow=2)
```

Look at a couple of the results as forest plots
```{r}
forestPlot(metaObj, "CYP3A4") 
forestPlot(metaObj, "ABCB1") 
```

Add the chromosome location + gene name
```{r}
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
pos.genes <- metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0$posGeneNames
neg.genes <- metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0$negGeneNames
sig.genes <- c(pos.genes, neg.genes)
mapping.table <- getBM(c('hgnc_symbol','chromosome_name', 'entrezgene'), "hgnc_symbol",sig.genes, mart=ensembl)
mapping.table2 <- mapping.table[mapping.table$chromosome_name %in% c(1:22, "X", "Y"),] # remove the two patch ones
table(mapping.table2$chromosome_name)
```

Plot effect size vs. chromosome!
```{r}
sigGenes <- rbind(metaRes$pos, metaRes$neg)
sigGenes <- sigGenes[!duplicated(sigGenes),]
sigGenes$hgnc_symbol <- rownames(sigGenes)
geneChrES <- full_join(select(sigGenes, effectSize, hgnc_symbol), mapping.table2)
geneChrES$chr_type <- sapply(geneChrES$chromosome_name, function(x) ifelse(x == "X", "X", ifelse(x=="Y", "Y", "A")))

# order by effect size for visualization
geneChrES2 <- geneChrES[order(geneChrES$effectSize),]
geneChrES2$hgnc_symbol <-factor(geneChrES2$hgnc_symbol, levels=unique(geneChrES2$hgnc_symbol[order(geneChrES2$effectSize)]))

ggplot(geneChrES2, aes(y=effectSize, x=hgnc_symbol, fill=chr_type))+ geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



Heatmaps using R package
```{r}
heatmapPlot(metaObj, metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0 )
heatmapPlot(metaObjv, metaObj$filterResults$FDR0.2_es0.3_nStudies2_looaTRUE_hetero0 )
```


Make a custom heatmap that displays gene names and includes discovery and validation and combined effect size.
```{r}
# format the effect size data from the discovery and validation
siggenesMA.filt <- sigGenes[sigGenes$numStudies >=8,]
geneStudyTab <- cbind(metaObj$metaAnalysis$datasetEffectSizes[rownames(siggenesMA.filt),], metaObjv$metaAnalysis$datasetEffectSizes[rownames(siggenesMA.filt),])
rownames(geneStudyTab) <- rownames(siggenesMA.filt)

colnames(geneStudyTab) <- c(names(metaObj$originalData), names(metaObjv$originalData))
short.beta <- siggenesMA.filt$effectSize # add the overall effect size

# put together and rotate
geneStudyTab2 <- cbind(geneStudyTab, short.beta)
colnames(geneStudyTab2)[16] <- c("combined")
rotTab <- t(as.matrix(geneStudyTab2))
new.order <- c(order(rowMeans(rotTab[1:15,])), 16)
rotTab2 <- rotTab[new.order,]
colOrder <- c(order(colMeans(rotTab[1:15,])))
rotTab3 <- rotTab2[,order(short.beta)]

# truncate the effect size for visualization purposes
rotTab3[rotTab3 > 1.5] <- 1.5
rotTab3[rotTab3 < -1.5] <- -1.5

# color the gene label by its chromosome location
siggenesMA.filt2 <- siggenesMA.filt
rownames(siggenesMA.filt2) <- siggenesMA.filt2$gene
siggenesMA.filt3 <- siggenesMA.filt2[colnames(rotTab3),]
chromLoc2 <- ifelse(siggenesMA.filt3$chr =="Y", "red", ifelse(siggenesMA.filt3$chr=="X","blue", ifelse(siggenesMA.filt3$chr=="XY", "purple", "black")))

# plot the heatmap
heatmap.2(rotTab3, trace="none", Rowv=FALSE, Colv=FALSE, colCol=chromLoc2, margins = c(5,9), dendrogram="none")

```



